---
title: "Get started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## What Syrona does

Syrona compares two datasets in the OMOP Common Data Model, for example two databases, two hospitals or two cohorts. For each dataset it counts how common every condition, procedure and drug is, per year, sex and age group. It then compares the two datasets with prevalence ratios and pools them with a meta-analysis. The results can be explored in a dashboard.

Syrona only reads the database. What it saves are aggregated counts, and no saved count is below 5.

## Install

```{r, eval=FALSE, runner=FALSE}
install.packages("syrona")

# Plus the driver for your database: one of
install.packages("RPostgres")  # a PostgreSQL server
install.packages("duckdb")     # a local DuckDB file
```

## Try it without a database

Syrona comes with two small demo datasets.

```{r, eval=FALSE}
library(syrona)

# Copy the demo datasets to a temporary folder
demo_dir <- file.path(tempdir(), "syrona_demo")
dir.create(demo_dir, showWarnings = FALSE)
file.copy(system.file("extdata", "demo", "data", package = "syrona"), demo_dir, recursive = TRUE)
options(syrona.data_dir = demo_dir)

# Compare them (overwrite = TRUE lets you run this again)
compare_all("demo_population", "demo_selected", overwrite = TRUE)

# Explore the result
run_app(data_dir = demo_dir)
```

## The whole path with your own data

Fill in the settings, then run the rest as it is. The example compares two CDM schemas on one PostgreSQL server.

```{r settings, eval=FALSE}
library(syrona)

# Settings (details: vignette("a01_connect"))
host         <- "localhost"
dbname       <- "omop"
user         <- "your_user"
cdm_schema_a <- "cdm_a"
cdm_schema_b <- "cdm_b"
data_dir     <- "path/to/syrona_output"  # results are saved here
options(syrona.data_dir = data_dir)
```

```{r, eval=FALSE, runner="postgres"}
# 1. Connect to each CDM (read-only)
db_a <- syrona_connect_pg(host = host, dbname = dbname, user = user, cdm_schema = cdm_schema_a)
db_b <- syrona_connect_pg(host = host, dbname = dbname, user = user, cdm_schema = cdm_schema_b)

# 2. Extract: counts per concept, year, sex and age group
extract_all("Dataset_A", db_a)
extract_all("Dataset_B", db_b)

# 3. Compare the two datasets
compare_all("Dataset_A", "Dataset_B")

# 4. Explore in the dashboard
run_app(data_dir = data_dir)

# 5. Disconnect
syrona_disconnect(db_a)
syrona_disconnect(db_b)
```

## Where the results are saved

```
path/to/syrona_output/
  data/
    sources/
      Dataset_A/                   # one folder per extracted dataset
        _metadata.csv
        condition_prevalence.csv   # counts per concept, year, sex and age group
        ...                        # the same for procedures and drugs
      Dataset_B/
    comparisons/
      Dataset_A_vs_Dataset_B/      # one folder per comparison
        condition_meta_summary.csv # one row per concept: the pooled result
        ...
```

## Next

- `vignette("a01_connect")`: connect to PostgreSQL or DuckDB, and check your access
- `vignette("a02_extract")`: extract a dataset
- `vignette("a03_populations")`: extract a cohort, a hospital or a list of patients
- `vignette("a04_compare")`: compare two datasets and open the dashboard
- `vignette("a05_methods")`: how the comparison is calculated
- `vignette("a06_vocabularies")`: how concepts are grouped into chapters
