---
title: "Connect"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Connect}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Syrona reads an OMOP CDM from PostgreSQL or from a DuckDB file. It only reads: it needs read access to the CDM schema and, on PostgreSQL, the right to create temporary tables.

## Settings

```{r settings, eval=FALSE}
library(syrona)

host       <- "localhost"  # the database server (localhost through an SSH tunnel)
port       <- 5432         # the port (with a tunnel: its local port)
dbname     <- "omop"       # the database
user       <- "your_user"  # your database user
password   <- Sys.getenv("DB_PASSWORD")  # from ~/.Renviron, see below
cdm_schema <- "cdm"        # the schema with the OMOP tables
```

The password is kept out of the script, in `~/.Renviron` (restart R after editing it):

```         
DB_PASSWORD=your_password
```

## PostgreSQL

```{r, eval=FALSE, runner="postgres"}
db <- syrona_connect_pg(
  host       = host,
  port       = port,
  dbname     = dbname,
  user       = user,
  password   = password,
  cdm_schema = cdm_schema
)
```

## DuckDB

The file is opened read-only.

```{r, eval=FALSE, runner="duckdb"}
# The OMOP tables are in the schema "main". If not, add cdm_schema = "..."
db <- syrona_connect("path/to/omop.duckdb")
```

## Check the connection

```{r, eval=FALSE}
# The OMOP tables Syrona found
names(db$cdm)

# The number of persons
db$cdm$person |> dplyr::tally()
```

On PostgreSQL, also check the right to create temporary tables:

```{r, eval=FALSE, runner="postgres"}
# Must be TRUE
DBI::dbGetQuery(db$con, "SELECT has_database_privilege(current_database(), 'TEMP') AS can_temp")
```

## Disconnect

```{r, eval=FALSE}
syrona_disconnect(db)
```

## If it does not connect

-   **"Connection refused"**: the SSH tunnel is not open, or `port` is not its local port.
-   **"password authentication failed"**: `DB_PASSWORD` is missing or wrong, or R was not restarted.
-   **"permission denied" or an error about temporary tables**: your user lacks a right to read the CDM schema or to create temporary tables (see "Check the connection"). Your database administrator can grant it.
