---
title: "Cohorts and hospitals"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cohorts and hospitals}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

By default `extract_all()` counts everyone in the database. It can also count a cohort, the patients of one hospital, or your own list of patients.

## Settings and connection

```{r settings, eval=FALSE}
library(syrona)

host           <- "localhost"
port           <- 5432
dbname         <- "omop"
user           <- "your_user"
cdm_schema     <- "cdm"
cohort_schema  <- "results"  # where ATLAS writes its cohorts
data_dir       <- "path/to/syrona_output"
options(syrona.data_dir = data_dir)
```

```{r, eval=FALSE, runner="postgres"}
db <- syrona_connect_pg(host = host, port = port, dbname = dbname, user = user,
                        password = Sys.getenv("DB_PASSWORD"), cdm_schema = cdm_schema)
```

```{r, eval=FALSE, runner="duckdb"}
db <- syrona_connect("path/to/omop.duckdb")
```

## A cohort generated in ATLAS

ATLAS writes its cohorts to the table `cohort` in the results schema. The cohort id is the ATLAS id, the name is shown in ATLAS.

```{r, eval=FALSE}
# Entries and persons of cohort 1
cohort_summary(db$con, cohort_id = 1, cohort_schema = cohort_schema)

# One dataset per cohort
extract_all("Dataset_A", db, cohort_id = 1, cohort_schema = cohort_schema)
extract_all("Dataset_B", db, cohort_id = 2, cohort_schema = cohort_schema)
```

A cohort table with another name is read with `cohort_table = "..."`.

## One hospital

A hospital dataset has the persons with at least one visit at that care site, and only the events recorded at those visits.

```{r, eval=FALSE, runner="duckdb"}
# The care sites with at least 100 patients (change with min_patients)
list_care_sites(db$con, cdm_schema = cdm_schema)

# One dataset per hospital
extract_all("Hospital_A", db, care_site_id = 101)
extract_all("Hospital_B", db, care_site_id = 102)
```

Events are linked to a hospital through their visit. Check first how many are linked, per domain:

```{r, eval=FALSE, runner="duckdb"}
# Share of drug events with a visit: close to 0 means no drug data per hospital
DBI::dbGetQuery(db$con, paste0("
  SELECT count(v.visit_occurrence_id) * 1.0 / count(*) AS linked
  FROM ", cdm_schema, ".drug_exposure d
  LEFT JOIN ", cdm_schema, ".visit_occurrence v ON d.visit_occurrence_id = v.visit_occurrence_id"))
```

## Your own cohort

Syrona reads any cohort table in the OHDSI format. To make your own, for example from ATLAS JSON files or from a list of patients, connect with a write schema: a schema where you may create tables, your own one (not shared with other users). Syrona adds the prefix `syrona_` to every table written there.

```{r, eval=FALSE}
syrona_disconnect(db)
```

```{r, eval=FALSE, runner=FALSE}
db <- syrona_connect_pg(host = host, port = port, dbname = dbname, user = user,
                        password = Sys.getenv("DB_PASSWORD"), cdm_schema = cdm_schema,
                        write_schema = "your_schema")
```

```{r, eval=FALSE, runner="duckdb"}
# A DuckDB file must then be opened writable
db <- syrona_connect("path/to/omop.duckdb", write_schema = "main", read_only = FALSE)
```

**From ATLAS JSON files.** Export the definitions from ATLAS into one folder. The ids follow the file names in alphabetical order. Generating needs Java and the packages CirceR and SqlRender: `install.packages(c("CirceR", "SqlRender"))`.

```{r, eval=FALSE, runner=FALSE}
cohort_set <- CDMConnector::readCohortSet("path/to/json_folder")
db$cdm <- CDMConnector::generateCohortSet(db$cdm, cohort_set, name = "atlas")
CDMConnector::settings(db$cdm$atlas)     # ids and names
CDMConnector::cohortCount(db$cdm$atlas)  # ids and counts

extract_all("Dataset_A", db, cohort_id = 1, cohort_table = "atlas")
```

**From a list of patients.** The ids must be OMOP `person_id`s. A wide date range keeps each person's whole observation period.

```{r, eval=FALSE, runner="duckdb"}
ids <- c(1, 2, 3, 4, 5)  # your person_ids
my_list <- data.frame(
  cohort_definition_id = 1L,
  subject_id = ids,
  cohort_start_date = as.Date("1900-01-01"),
  cohort_end_date = as.Date("2100-12-31")
)
db$cdm <- CDMConnector::insertTable(db$cdm, name = "my_list", table = my_list)

extract_all("My_patients", db, cohort_id = 1, cohort_table = "my_list")
```

```{r, eval=FALSE}
syrona_disconnect(db)
```

## Rules

- Say where a cohort is: `cohort_schema`, `cohort_table` or both. Syrona does not look for a table on its own.
- A person's entries in a cohort must not overlap. Cohorts generated in ATLAS never do.
- `cohort_id` and `care_site_id` cannot be combined.
- The cohort or care site used is recorded in `_metadata.csv`. Counts below 5 are written as `<5`.

## Coming from Syrona 0.2.1

- `insert_cohort()` is removed: use a cohort in the database, or make your own as above.
- `create_caresite_cohort()` and `delete_cohort()` are removed: use `care_site_id`. Results differ from 0.2.1, which counted all events between a person's first and last visit.
- `extract_all(cohort_id = , cohort_schema = )` works as before.
