---
title: "OHDSI Cohort Workflows"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{a03_cohorts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

By default, `extract_all()` extracts data for all persons in the database. But often you want to compare specific subpopulations - patients from a particular hospital, a clinical trial cohort, or a phenotype definition.

Syrona supports OHDSI-standard cohort tables for this. A cohort table has 4 columns:

| Column                 | Type    | Description     |
|------------------------|---------|-----------------|
| `cohort_definition_id` | INTEGER | Which cohort    |
| `subject_id`           | INTEGER | person_id       |
| `cohort_start_date`    | DATE    | When they enter |
| `cohort_end_date`      | DATE    | When they exit  |

When you pass a `cohort_id` to `extract_all()`, it restricts extraction to cohort members.

## Defining a cohort

Following the OHDSI convention, a cohort is *"a set of persons who satisfy one or more inclusion criteria for a duration of time"* ([The Book of OHDSI, Cohorts](https://ohdsi.github.io/TheBookOfOhdsi/Cohorts.html)). Each person's membership is bounded by a **cohort window**: `cohort_start_date` (entry / index date) to `cohort_end_date` (exit - typically the end of a persistence window or of the observation period).

Syrona does not author rule-based phenotype definitions itself; it **reads** any OHDSI-standard cohort table, for example a rule-based definition from ATLAS or [Capr](https://ohdsi.github.io/Capr/), or a cohort built with [CohortConstructor](https://ohdsi.github.io/CohortConstructor/). As long as the result is the standard four-column cohort table, `extract_all(cohort_id = ...)` can use it. Comparing hospitals needs no cohort table at all (next section).

## Hospitals (care sites)

The most common use case: compare hospitals within the same OMOP CDM database. `extract_all(care_site_id = )` extracts one care site. Nothing is written to the database.

```{r, eval=FALSE}
library(syrona)

cdm_schema <- "cdm"   # change to your CDM schema

db <- syrona_connect_pg(
  dbname = "omop",
  user = "analyst",
  cdm_schema = cdm_schema
)

# Find available care sites
list_care_sites(db$con, cdm_schema = cdm_schema)
#>   care_site_id    care_site_name  n_patients
#> 1          101   Central Hospital      45000
#> 2          205   University Clinic     28000
#> 3          312   Regional Hospital     15000

# Extract each hospital's data, then compare
extract_all("Central_Hospital", db = db, care_site_id = 101)
extract_all("University_Clinic", db = db, care_site_id = 205)
compare_all("Central_Hospital", "University_Clinic")
```

### What a care-site dataset contains

-   **Persons**: everyone with at least one visit at the care site, over their whole observation period (this is the denominator)
-   **Events**: only conditions, procedures and drugs recorded at a visit at that care site, linked through `visit_occurrence_id`

A person who visited two hospitals is counted in both datasets, each time with only the events recorded at that hospital. This is the method of the published Syrona analysis.

### Check the visit links first

Events with no `visit_occurrence_id`, or one that points to no visit, cannot be assigned to a care site and are left out. In some databases this is common, for example drugs in claims data. `extract_all()` shows a note for every domain with no linked events. To see the share of linked events before extracting:

```{r, eval=FALSE}
DBI::dbGetQuery(db$con, "
  SELECT count(v.visit_occurrence_id) * 1.0 / count(*) AS linked_share
  FROM cdm.drug_exposure d
  LEFT JOIN cdm.visit_occurrence v
    ON d.visit_occurrence_id = v.visit_occurrence_id")
```

Replace `cdm` with your CDM schema, and repeat with `condition_occurrence` and `procedure_occurrence`. A share close to 0 means that domain will be empty or nearly empty for every care site.
