---
title: "Compare and explore"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Compare and explore}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`compare_all()` compares two extracted datasets, see `vignette("a02_extract")`. It needs no database connection.

## Settings

```{r settings, eval=FALSE}
library(syrona)

data_dir <- getwd()  # the folder with your extracted datasets (the one used when extracting)
options(syrona.data_dir = data_dir)
```

## Compare

```{r, eval=FALSE}
compare_all("Dataset_A", "Dataset_B")
```

The result is saved in `data/comparisons/Dataset_A_vs_Dataset_B/`. To compare the same two datasets again, add `overwrite = TRUE`.

For each concept, `log2_pr` is the log2 prevalence ratio of Dataset_B to Dataset_A: 1 means twice as common in Dataset_B, -1 half as common. The final result per concept is in `condition_meta_summary` (and `procedure_`, `drug_`). How it is calculated: `vignette("a05_methods")`.

## Explore

```{r, eval=FALSE}
run_app(data_dir = data_dir)
```

If the dashboard lists no comparisons, `data_dir` is not the folder that contains `data/comparisons/`.
